Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-31 paper/main.tex
more comments.

@@

\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Black-box combinatorial optimization problems arise in practice when the
  objective function is evaluated by means of a simulator or a real-world
  experiment. In such cases, classical techniques such as mixed-integer
  programming and local search cannot be applied. Moreover, often each solution
  evaluation is expensive in terms of time or resources, thus only a limited
  number of evaluations is possible, typically several order of magnitude
  smaller than in white-box optimization problems. In the continuous case,
  Bayesian optimization, in particular using Gaussian processes, has proven
  very effective under these conditions. Much less research is available in the
  combinatorial case. In this paper, we propose and analyze an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and compare it with CEGO, a Bayesian optimization algorithm for
  combinatorial optimization. Experimental results show that the Bayesian
  algorithm is able to obtain very good solutions with very few evaluations,
  however, at a significant computational cost, whereas the proposed EDA
  outperforms CEGO when the number of solutions evaluated approaches 400, and
  it is significantly faster. These results suggest that the combination of
  Bayesian optimization and a Mallows-based EDA may be an interesting direction
  for future research.\sloppy  
\textcolor{red}{16 pages max, deadline: 1 November 2020}
\keywords{Combinatorial optimization \and Bayesian optimization \and Expensive black-box optimization \and Estimation of distribution algorithms}
\end{abstract}

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}, such as Kendall's $\tau$ distance for
permutations (Eq.~\ref{eq:kendall_decomp}). In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. The original paper notes that coupling the GA
with local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness
function~\citep[p.~875]{ZaeStoFriFisNauBar2014}.

@@

\subsubsection{CEGO}
%
We use the implementation of GECO provided by the
authors.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of $\FEmax=400$
true fitness evaluations requires \emph{more than a week} of computation time,
which was infeasible for our study. Hence, we set the budget of the GA to
$10^3$. Although a higher GA budget could improve the results of CEGO reported
here, our results are similar (and sometimes better) than the ones reported in
the original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. In fact, we found that the
increase in runtime caused by doubling $\FEmax$ is much smaller than by increasing
the GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as whitespace.  Results for all the datasets can be found in the supplemental material.

\begin{figure}[htb]
  \centering%
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d100.01} (LOP)\\[-0.2ex]
    \includegraphics[width=\textwidth]{../img/heatmap_N-t1d100_01}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d50.01} (LOP)\\[-0.2ex]
 \includegraphics[width=\textwidth]{../img/heatmap_N-t1d150_01}
\end{minipage}\\\vskip 1ex
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_rec05}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{kra32} (QAP)\\[-0.5ex]
\includegraphics[width=\textwidth]{../img/heatmap_kra32_dat}
\end{minipage}
\caption{Mean fitness (over 10 runs) of different configurations of $r_1$ and $r_2$ (darker is better). Whitespace denotes missing data (runs that took longer than 6 hours).\label{fig:heatmaps}}
\end{figure}

The plot shows that UMM is indeed very sensitive to the values of $r_1$ and
$r_2$, yet, low values of $r_1$ and high values of $r_2$ almost always produce
the best results, independently of the problem. Nevertheless, there are a few
cases where an interaction seems to occur and particular combinations perform
remarkably well.\MANUEL{Missing data, whitespaces where more than 6 hours of wall-clock time}  In particular, we observe that the pattern on
the two left-most figures differs from the two right-most ones. While on the
left the fitness gets worse homogeneously from the top right corner to the
bottom left one, the pattern on the right is not as homogeneous, yet still
fitness gets worse with decreasing $r_2$.  We observe that this behavior is
related to the adequacy of UMM to the problem: the homogeneous behavior of
parameters $r_1$ and $r_2$ depends on the problem (not to the particular
instances) and the more homogeneous the behavior, the better performance of
UMM. We elaborate this argument in the following sections with detail.\MANUEL{top and bottom}

Over all the instances  we
see that the configuration $r_1 =0.1$, $r_2=0.9$ has good
performance: \ekhine{}{the rank of this parameter configuration among all the possible configurations is less than 4 on average for all the instances}. From this
preliminary experiment, it is not obvious how the best settings are related to
problem features, such as permutation length ($n$). On the other hand, given
the sharp transitions in the plots, it is clear that further fine-tuning may
improve performance and uncover further patterns.  Although automatic parameter
tuning may likely provide even more fine-tuned settings, our goal here is to
understand how UMM performs in comparison to an existing Bayesian method (CEGO)
and automatically tuning the parameters of CEGO is not computationally
feasible.  For consistency, in the following sections, we run UMM with this
parameter configuration. Moreover, we recommend this configuration as the
current default of UMM.

\subsection{Analysis of UMM and CEGO}

@@

When we compare the behavior of CEGO and
UMM on synthetic LOP instances (Fig.~\ref{fig:lop_synth}), we observe a clear
pattern independently of the uniformity of the instances. In particular, we see
a linear convergence to the optimal value for UMM.  This is due to the linear
decrease in the parameter $\theta$ (line~\ref{line:theta} in
Algorithm~\ref{alg:umm}), which imposes a linear variance decrease on the
internal probability model of the algorithm over the $m$ functions
evaluations. Overall, \ekhine{UMM found  the best solutions in all the repetitions. }{the mean best solutions of UMM are better than the mean of CEGO regardless the uniformity of the instances. }

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values (e.g., $\phi=.5$) show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}

\newcommand{\supplement}{\citep{Supplementary}}

@@

  In the following paragraphs we analyze the result over real rata. Fig.~\ref{fig:loplib} shows the results on the instances available in LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
  The distribution of this instances differs from the symthetic LOP instances, and thus so does the landscape. However, UMM shows a very similar performance: there is a (close to) linear decrease on the function evaluations of new solutions along iterations. This suggests that UMM very good fits the fitness landscape. Regarding CEGO, we can take similar conclussions for the first 50-100 iterations. However, the processes had to be halted after hitting the wall time, so no conlusion can be taken for further iterations. Overall, the improvement from the first (random) to the last solutions is 2-3 times larger in UMM than CEGO.

In the case of the PFSP, none of the algorithms are able to achieve a progress
better than a random search would, as illustrated by the results
(Fig.~\ref{fig:rec05}) obtained on \texttt{rec05}, which is the smallest PFSP
instance considered here. The lack of any apparent convergence suggests that
the models (both Bayesian and probabilistic) are not learning anything about
the fitness landscape. Results for larger PFSP corroborate these conclusions
(\supplement). Although previous studies~\citep{ZaeStoBar2014:ppsn} concluded
that other distance measures are more suited than Kendall's-$\tau$ distance for
the PFSP, the reported fitness differences between various distance metrics are
small, hence, we conjecture that the same behaviour will be observed with other
distance metrics.

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra32}. As shown in Fig~\ref{fig:kra32}, in this instance both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet, around evaluation 250, it appears to
get stuck and around evaluation 300, UMM overtakes it. We believe the reason is
the same as before, that is, the model is not able to adequately reflect the
ruggedness of the actual landscape and, hence, the best solutions found by the
GA searching the model are not improved solutions for the actual problem. Interestingly, \texttt{kra32} is rectangular
whereas \texttt{tho32} contains rectangular distances and \texttt{nug12} and \texttt{nug30} contain Manhattan distances of rectangular grids.\MANUEL{so they all have the same distance type!!! What is the difference then? Ask Thomas?}

@@

Table~\ref{tab:results} shows overall results over all instances evaluated (excluding the synthetic LOP instances). In particular, the mean fitness (and standard deviation) of the best solution found at the end of each run,
In terms of best fitness achieved on each run, and taking into account the
random behaviour observed in some instances and discussed above,\ekhine{}{falta un cacho aqui}\MANUEL{Mío}

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run went over 72 hours, of  CEGO was not able to reach 400 function evaluations (FEs) in 72 hours, it was terminated earlier. This was the case for all LOP instances shown.\manuel{}{Don't we have results for rec31?}\label{tab:results}}

@@

    QAP     & kra32        & 115907.0                & (4663.4)   & 116320.0  & (4673.5)               & $[$-4799.3,   & 3973.3$]$     & 400.0& 95.8 & 0.3  \\
        & nug12        & 624.2                   & (26.3)     & 660.4     & (17.3)                 & $[$-57.3,     & -15.1$]$      & 400.0& 14.8 & 0.0  \\

@@

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer\EKHINE{hhmmm lo hacemos mejor o igual}. The results presented here
show that, despite the intrinsic difficulty of this scenario, there are still
significant advances yet to be made. Thus, we hope that this paper will
motivate further research. % on such problems.

>>>>

\newcommand{\supplement}{~(\href{http://doi.org/10.5281/zenodo.4167148}{doi:10.5281/zenodo.4167148})}

@@

\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Black-box combinatorial optimization problems arise in practice when the
  objective function is evaluated by means of a simulator or a real-world
  experiment. In such cases, classical techniques such as mixed-integer
  programming and local search cannot be applied. Moreover, each fitness
  evaluation is often expensive in terms of time or resources, thus only a
  limited number of evaluations is possible, typically several orders of
  magnitude smaller than in white-box optimization problems. In the continuous
  case, Bayesian optimization, in particular using Gaussian processes, has
  proven very effective under these conditions. Much less research is available
  in the combinatorial case. In this paper, we propose and analyze UMM, an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and unbalanced rank aggregation (uBorda). Experimental results on
  black-box versions of LOP, QAP and PFSP show that UMM is able to match, and
  sometimes surpass, the solutions obtained by CEGO, a Bayesian optimization
  algorithm for combinatorial optimization, specially when the budget of
  solutions evaluated approaches 400. The computational cost of UMM is at least
  an order of magnitude smaller than CEGO, which makes it specially useful when
  both computation time and number of evaluations are of concern.\sloppy
  \textcolor{red}{16 pages max, deadline: 1 November 2020}
  \keywords{Combinatorial optimization \and Bayesian optimization \and
    Expensive black-box optimization \and Estimation of distribution
    algorithms}
\end{abstract}

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}, such as Kendall's $\tau$ distance for
permutations (Eq.~\ref{eq:kendall_decomp}). In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. The original paper notes that coupling the GA
with local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness
function~\citep[p.~875]{ZaeStoFriFisNauBar2014}. A thorough comparison of CEGO
with other optimizers for expensive black-box combinatorial optimization ranked
it as the best-performing~\citep{ZaeStoFriFisNauBar2014}.

@@

\subsubsection{CEGO}
%
We use the implementation of GECO provided by the
authors.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of
$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
computation time, which was infeasible for our study. Due to extremely long
runtimes of CEGO and constraints in our computing system, we set a maximum
wall-clock time of 3 days for each run of CEGO. Hence, we set the budget of the
GA to $10^3$. Although a higher GA budget could improve the results of CEGO
reported here, our results are similar (and sometimes better) than the ones
reported in the original
papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to running CEGO
with $\FEmax=400$ instead of $\FEmax=200$. In fact, we found that the increase
in runtime caused by doubling $\FEmax$ is much smaller than by increasing the
GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as white.  Results for all the datasets can be found in the supplemental material\supplement.

\begin{figure}[htb]
  \centering%
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d100.01} (LOP)\\[-0.2ex]
    \includegraphics[width=\textwidth]{../img/heatmap_N-t1d100_01}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d50.01} (LOP)\\[-0.2ex]
 \includegraphics[width=\textwidth]{../img/heatmap_N-t1d150_01}
\end{minipage}\\\vskip 1em
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_rec05}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{kra32} (QAP)\\[-0.5ex]
\includegraphics[width=\textwidth]{../img/heatmap_kra32_dat}
\end{minipage}
\caption{Mean fitness (over 10 runs) of different configurations of $r_1$ and $r_2$ (darker is better). Whitespace denotes missing data (runs that took longer than 6 hours).\label{fig:heatmaps}}
\end{figure}

The plot shows that UMM is indeed very sensitive to the values of $r_1$ and
$r_2$, yet, low values of $r_1$ and high values of $r_2$ almost always produce
the best results, independently of the problem.  In particular, we observe that
the pattern on the two top figures differs from the two bottom ones. While on
the top figures the fitness gets worse homogeneously from the top-right corner
to the bottom-left one, the pattern on the bottom figures is not as homogeneous, yet
still fitness gets worse with decreasing $r_2$.  We observe that this behavior
is related to the adequacy of UMM to the problem: the homogeneous behavior of
parameters $r_1$ and $r_2$ depends on the problem (not on the particular
instances) and the more homogeneous the behavior, the better performance of
UMM. We elaborate this argument in detail in the next section.

From this preliminary experiment, it is not obvious how the best settings are
related to problem features, such as permutation length ($n$). On the other
hand, given the sharp transitions in the plots, it is clear that further
fine-tuning may improve performance and uncover further patterns.

Over all the instances, we observe that the configuration $r_1 =0.1$, $r_2=0.9$
has good performance, i.e., its mean rank (over 10 runs) relative to all the
other configurations is less than 4 for all the instances.  Although automatic
parameter tuning may likely provide even more fine-tuned settings, our goal
here is to understand how UMM performs in comparison to an existing Bayesian
method (CEGO) and automatically tuning the parameters of CEGO is not
computationally feasible.  For consistency, in the following sections, we run
UMM with this parameter configuration.  Moreover, we recommend this
configuration as the current default of UMM.

\subsection{Analysis of UMM and CEGO on synthetic LOP instances}

@@

When we compare the behavior of CEGO and
UMM on synthetic LOP instances (Fig.~\ref{fig:lop_synth}), we observe a clear
pattern independently of the uniformity of the instances. In particular, we see
a linear convergence to the optimal value for UMM.  This is due to the linear
decrease in the parameter $\theta$ (line~\ref{line:theta} in
Algorithm~\ref{alg:umm}), which imposes a linear variance decrease on the
internal probability model of the algorithm over the $\FEmax$ functions
evaluations. Overall, the mean fitness returned by UMM after $\FEmax$ is better one returned by  CEGO, regardless the uniformity of the instances.

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values (e.g., \manuel{$\phi=.5$}{0.5})\MANUEL{We never use $\phi$, not even in the plots, so less have one less symbol to worry about.} show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}

@@

\subsection{Analysis of UMM and CEGO on LOP, QAP and PFSP}

In the following paragraphs we analyze the result over real-world instances. %
Figure~\ref{fig:loplib} shows the results on the instances taken from LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
The distribution\MANUEL{what do you mean the distribution? Is this a feature of
  the matrix $a$?} of these instances differs from the synthetic LOP instances,
and, thus, so does the landscape. However, UMM shows a very similar behavior:
there is a (close to) linear decrease \manuel{on the function evaluations of
  new solutions along iterations}{of the fitness of new solutions over each new
  evaluation}. This suggests that UMM \manuel{very good fits}{is able to model}
the fitness landscape. Regarding CEGO, we \manuel{can take similar
  conclusions}{observe a similar behavior} for the first 50-100
\manuel{iterations}{evaluations}. At this point, CEGO reaches the time limit of
3 days, so no observations can be made for longer budgets.  Overall, the
improvement from the first (random) to the last solutions is 2-3 times larger
in UMM than CEGO. %
\MANUEL{But this is a bit misleading because UMM runs for 4 times more
  evaluations, perhaps we should also measure up to the evaluation that CEGO
  reaches so we report both values?}

In the case of the PFSP, none of the algorithms are able to achieve a progress
better than a random search would, as illustrated by the results
 obtained on \texttt{rec05} (Fig.~\ref{fig:rec05}), which is the smallest PFSP
instance considered here. The lack of any apparent convergence suggests that
the models (both Bayesian and probabilistic) are not learning anything about
the fitness landscape. Results for larger PFSP corroborate these conclusions\supplement.
\manuel{Although previous studies~\citep{ZaeStoBar2014:ppsn} concluded
that other distance measures are more suited than Kendall's-$\tau$ distance for
the PFSP, the reported fitness differences between various distance metrics are
small.
Hence, we conjecture that the same behavior will be observed with other distance metrics.}{}\MANUEL{On second thought, I'm not sure we should say this, because we really have no idea and we actually hope it is not true.}

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra32}. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to
get stuck around evaluation 250,  and UMM overtakes it around evaluation 300. We believe the reason is
the same as before, that is, the CEGO model is not able to adequately reflect the
ruggedness of the actual landscape and, hence, the best solutions found by the
GA searching the model are not improved solutions for the actual problem. Interestingly, \texttt{kra32} is rectangular
whereas \texttt{tho32} contains rectangular distances and \texttt{nug12} and \texttt{nug30} contain Manhattan distances of rectangular grids.\MANUEL{so they all have the same distance type!!! What is the difference then? Ask Thomas?}

@@

Table~\ref{tab:results} shows results over all instances evaluated (excluding
the synthetic LOP instances). In particular, we show the mean fitness (and
standard deviation) of the best solution found at the end of each run. We also report the mean runtime in hours and, since CEGO sometimes reached the limit of 72 hours, also the number of fitness evaluations reached by CEGO.

In the LOP, differences are mostly in favour of UMM, due to the time limit of
72 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 16 hours. If UMM was allowed to reach the 72 hours limit, the differences
would be even stronger.

In the case of the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search. Similar observations can be made for the QAP, where
only on the smallest instance \texttt{nug12}, there is a significant different
in favour of CEGO.

The overall conclusion from Table~\ref{tab:results} is that there is not a
significant difference in performance between UMM and CEGO. For the larger LOP
instances, CEGO reaches the time limit of 72 hours before consuming the budget
of 400 evaluations and thus, obtains quite poor solutions compared to UMM,
which is at least an order of magnitude faster.

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run of  CEGO was not able to reach 400 function evaluations (FEs) in 72 hours, it was terminated earlier. This was the case for all LOP instances shown.\manuel{}{Don't we have results for rec31?}\label{tab:results}}

@@

    QAP  & nug12        & 624.2                   & (26.3)     & 660.4     & (17.3)                 & $[$-57.3,     & -15.1$]$      & 400.0& 14.8 & 0.0  \\

@@

         & kra32        & 115907.0                & (4663.4)   & 116320.0  & (4673.5)               & $[$-4799.3,   & 3973.3$]$     & 400.0& 95.8 & 0.3  \\

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
\MANUEL{Pero UMM es peor en rec13 y nug12, y en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir lo que pasa en el momento en que matamos a CEGO y dar también ese valor entonces la diferencia en velocidad no importaría. }
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here
show that, despite the intrinsic difficulty of this scenario, there are still
significant advances yet to be made. Thus, we hope that this paper will
motivate further research. % on such problems.
